#Introduction to base R - 7th October 2019

R Studio - interactive development environment

Tools > Global options > Git/SVN > Type in Git details > Enable version control interface for R Studio projects
Should then get Git tab on top right for commiting etc.
New files will appear in environment, tick to add then press commit. Can type commit message. Can push separately once 
set up with online repository.
Can also set up project by cloning from Git repository (eg template files)

Packages for R: CRAN (package repository, general use), Bioconductor (bioinformatics, more documentation/vignettes)
Github also has some packages but not as rigorously tested as CRAN or Bioconductor, so be wary.

Install via Conda on command line if want to keep track of package versions etc for example:
$conda install r-tidyverse
$conda install bioconductor-deseq2

Better to use <- notation not = for assigning variables. (alt + - as a shortcut)

Attributes associated with each value eg names or other metadata - similar to python dictionaries
>attributes(vector_name)

R is nested functions, can use tidyverse to make code look nicer e.g tibbles are tidyverse dataframes (simple and better functionality)

Good resourses:
R for datascience https://r4ds.had.co.nz/ 
Bioconductor 2018 Workshop https://bioconductor.github.io/BiocWorkshops/r-and-bioconductor-for-everyone-an-introduction.html
Short tutorial/refresher https://bioinformatics-core-shared-training.github.io/r-for-medics/notes.nb.html


```{r examples}
df[1] #1st column only
df[1,] #1st row only
df[5:10,1:2] #rows 5:10, columns 1:2

df[grep("TSS", df&=$column1),] #get rows containing TSS in column 1

df[df$column1 %in% c('promoter', 'enhancer'),] #rows that are enhancer or promoter values in column 1
df[df$column1 == 'promoter' | df$column1 == 'enhancer',] #rows that are enhancer or promoter values in column 1

colnames(table_name)[-c(1:2)] #remove first two columns (but only shows column names, need to put within table name to get full table)

#for loop in R
for (variable in sequence){
  Do something
}

if (i>3){
  print('Yes')
} else {
  print('No')
}

#apply methods
lapply(y,log) #list apply, do log of all values in y and return as list

getwd() #get working directory
setwd() #set working directory
list.files() #list files in directory
system('ls|tail') #run command line functions in R


MCF_down_cell_cycle<-MCF_down[is.element(MCF_down$SYMBOL, cell_cycle),] #look up values from one table (cell cycle) in another (MCF_down) to get all information from row

```
Summary of exercises in slack

```{r activity_1}

vector_1 <- 1:200
vector_1 * 123
vector_1[44]
vector_2 <- vector_1[1:15]
x <-  vector_2
mean(x)
median(x)
vector_2 <- c('actb', 100, 3.4) #items in vector are character type
number<-as.numeric(vector_2[2])
number*4

```

```{r activity_2}
sample_info <-  read.csv("sample_info.csv", header=TRUE, sep=',')
df <- as.data.frame(sample_info)
dim(df)
row_5 <- df[5,]
entry_4 <- df[4, "sample_title"]
df$mill_reads <- df$base_count/10000000
barplot(df$mill_reads, names.arg=df$sample_title, cex.names = 0.5, las=2) #simple barplot, adjust font size and make x labels vertical

```

```{r activity_3}
read_counts <- read.csv("read_counts.csv", header=TRUE, sep=',')
summary(read_counts)
counts <- read_counts
counts$X <- NULL #remove column X
counts <- counts[colnames(counts)[-c(2, 3, 4, 5, 6)]] #remove 2nd, 3rd, 4th, 5th and 6th columns
rownames(counts)<- counts$Geneid
counts$Geneid <-  NULL
counts$total_gene_count <- rowSums(counts)
new_counts <- counts[counts$total_gene_count!=0,]
sorted_counts <- new_counts[order(-new_counts$total_gene_count),]
high_genes <- head(sorted_counts, n=10)
low_genes <- tail (sorted_counts, n=10)

```
